Using Google Community Mobility Reports to investigate the growth of COVID-19 in the United States

### 2020-05-11 Covid-19-Google-CMR-US/Covid-19-Google-CMR-US.tex
Updated link to repository
Second version of paper after first round of anonymous reviews.

  \begin{abstract}
  In 2020 Google released a set of Community Mobility Reports (GCMR).
  These reports are based on the company's location-tracking capabilities
  and measure changes in mobility with respect to a baseline. This novel
  source of data offers an opportunity to investigate the way mobility
  potentially correlates with transmission of COVID-19. Using data from
  the New York Times on COVID-19 cases and GCMR, this paper presents an
  analysis of mobility levels and new daily reported cases of COVID-19 by
  state in the US. The results provide insights about the utility and
  interpretability of GCMR for COVID-19 research and decision-making.
  \end{abstract}

@@

The source document for this paper is an \texttt{R} markdown document
available from the following repository:

https://github.com/paezha/Google-Mobility-Reports-and-COVID-19-US

@@

GCMR use aggregated and anonymized data to chart changes in mobility
with respect to different classes of places, namely retail and
recreation, groceries and pharmacies, parks, transit stations,
workplaces, and residential. Mobility indicators for each of these
places are reported as percentage changes from a baseline level, which
corresponds to the median value of mobility of identical days of the
week during the period between January 3 and Feb 6, 2020. Covid-19 data
is compiled by The New York Times based on reports from state and local
health agencies.

For analysis, all mobility indicators are centered so that the value of
1 is the base mobility, and a 0.01 deviation corresponds to a 1\%
change. The incubation time of the disease has been estimated to be
between 2 and 12 days (95\% interval) by Lauer et al.~(2020). Given
this, it is to be expected that any changes in mobility will have a
lagged effect on the discovery of new cases. For this reason, lagged
moving averages of the mobility indicators are calculated. Furthermore,
it is possible that mobility and reports of new cases of COVID-19 are
endogenous, if the public adjust their mobility according to reports of
the incidence. Therefore, in addition to being consistent with an
incubation period, use of lagged indicators also helps to break this
potential endogeneity.

\begin{table}[!h]

\caption{\label{tab:descriptive-statistics}\label{tab:descriptive-statistics}Descriptive statistics of the data set}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{llllll}
\toprule
Variable & Definition & min & median & max & sd\\
\midrule
\rowcolor{gray!6}  New Cases & New Daily Cases of COVID-19 & 0 & 77.5 & 12126 & 1059.81\\
date & Date & 2020-02-27 & 2020-04-01 & 2020-04-26 & \\
\rowcolor{gray!6}  retail & Retail and Recreation & 0.34 & 0.66 & 1.16 & 0.22\\
groceries & Groceries and Pharmacies & 0.66 & 0.95 & 1.26 & 0.13\\
\rowcolor{gray!6}  parks & Parks & 0.36 & 1.15 & 1.8 & 0.24\\
\addlinespace
transit & Transit Stations & 0.24 & 0.73 & 1.14 & 0.24\\
\rowcolor{gray!6}  work & Workplaces & 0.34 & 0.65 & 1.07 & 0.2\\
residential & Residential & 0.97 & 1.14 & 1.27 & 0.08\\
\bottomrule
\multicolumn{6}{l}{\textit{Note: }}\\
\multicolumn{6}{l}{All mobility indicators are lagged 11-day moving averages}\\
\end{tabular}}
\end{table}

The lagged indicators are calculated as the mean of the mobility
indicator using the values from date-minus-12-days to date-minus-2-days.
Furthermore, using the cumulative number of reported COVID-19 cases, the
total number of new daily cases is calculated. This variable
(log-transformed after adding a small constant) is paired to the
corresponding lagged moving average of the mobility indicators. The
log-transformation is useful to avoid negative values of daily new cases
when making predictions. Analysis is based on correlation analysis,
multivariate regression, and data visualization. Table
\ref{tab:descriptive-statistics} shows the descriptive statistics of the
data set. Temporal coverage is from February 27 to April 26, 2020. The
maximum number of new daily cases during this period is 12,126.

@@

Table \ref{tab:correlation-analysis} shows that the mobility indicators
are highly correlated with each other. This is not surprising: it is
well-known that mobility involves time-use trade-offs: the more there is
of one kind of mobility, the less time there is available for any other.
That said, the strongest correlation with the outcome of interest is for
residential-based mobility. To avoid multicollinearity, parks-related
mobility is selected as a covariate since it has the lowest correlation
with residential-based mobility.

\begin{table}[!h]

\caption{\label{tab:check-correlations}\label{tab:correlation-analysis}Simple correlation between log(New Cases) and the mobility indicators}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lrrrrrrr}
\toprule
  & log\_new\_cases & retail & groceries & parks & transit & work & residential\\
\midrule
\rowcolor{gray!6}  log\_new\_cases & 1.00 & -0.68 & -0.47 & -0.30 & -0.68 & -0.69 & 0.71\\
retail & -0.68 & 1.00 & 0.86 & 0.44 & 0.93 & 0.98 & -0.98\\
\rowcolor{gray!6}  groceries & -0.47 & 0.86 & 1.00 & 0.46 & 0.87 & 0.87 & -0.86\\
parks & -0.30 & 0.44 & 0.46 & 1.00 & 0.52 & 0.44 & -0.45\\
\rowcolor{gray!6}  transit & -0.68 & 0.93 & 0.87 & 0.52 & 1.00 & 0.94 & -0.95\\
\addlinespace
work & -0.69 & 0.98 & 0.87 & 0.44 & 0.94 & 1.00 & -0.99\\
\rowcolor{gray!6}  residential & 0.71 & -0.98 & -0.86 & -0.45 & -0.95 & -0.99 & 1.00\\
\bottomrule
\multicolumn{8}{l}{\textit{Note: }}\\
\multicolumn{8}{l}{All mobility indicators are lagged 11-day moving averages}\\
\end{tabular}}
\end{table}

A regression model uses the log of new daily cases as the dependent
variable. The covariates are parks- and residential-related mobility,
which enter the regression in the form of a second order polynomial
expansion. In addition, the date is introduced to account for the
temporal trend of the pandemic. Finally, an indicator variable for the
state of New York and an interaction with the date are used to
distinguish the unusually high incidence of the disease there. The
results appear in Table \ref{tab:model-results}. The model provides a
good fit to the data and all variables reported are significant at
\(p<0.05\) or better.

There is an overall temporal trend that indicates a growing number of
cases over time, but at a decreasing rate (see negative sign of
date\^{}2). New York has on average more new daily cases than the rest
of the states, but this has also declined over time (see negative sign
of NY x date). Visualization is the most effective way to understand the
trend according to the mobility indicators and date. Figure
\ref{fig:prediction-plots} shows the prediction surfaces on three
different dates: March 21, when the first states began implementing
stay-at-home orders; April 5, fifteen days later; and April 20, fifteen
more days later, and at a time when some states are considering relaxing
stay-at-home orders. On March 21 there were still only minor departures
from baseline mobility (recall that these are temporally lagged); the
prediction surface at this point is relatively flat. This changes by
April 5, when large increases in residential-based mobility are seen,
along with greater variations in parks-based mobility. The prediction
surface indicates an expectation of a greater number of new daily cases
as both classes of mobility increase. On the last date examined, April
20, the trend becomes more steep for park-based mobility, even as this
indicator continues to display large variations from the baseline in
both directions. In general, higher levels of mobility tend to be
associates with a higher number of new daily cases.

\begin{table}[!h]

\caption{\label{tab:model-results}\label{tab:model-results}Results of estimating regression model. Dependent variable is log(New Cases + 0.0001).}
\centering
\begin{tabular}[t]{lcc}
\toprule
Variable & Coefficient Estimate & p-value\\
\midrule
\rowcolor{gray!6}  date & 7.9619 & <0.0001\\
date\textasciicircum{}2 & -0.0411 & <0.0001\\
\rowcolor{gray!6}  parks\textasciicircum{}2 & 1.4017 & 0.0251\\
parks & -37.2492 & <0.0001\\
\rowcolor{gray!6}  parks x residential & 39.0902 & <0.0001\\
\addlinespace
residential & -286.7568 & <0.0001\\
\rowcolor{gray!6}  residential\textasciicircum{}2 & 249.9901 & <0.0001\\
parks x date & -0.1412 & <0.0001\\
\rowcolor{gray!6}  residential x date & -6.9180 & <0.0001\\
residential x date\textasciicircum{}2 & 0.0365 & <0.0001\\
\addlinespace
\rowcolor{gray!6}  NY & 6.9864 & <0.0001\\
NY x date & -0.0557 & 0.0022\\
\bottomrule
\multicolumn{3}{l}{\textit{Note: }}\\
\multicolumn{3}{l}{Coefficient of Determination $R^2$= 0.821}\\
\multicolumn{3}{l}{Adjusted Coefficient of Determination $R^2$= 0.82}\\
\multicolumn{3}{l}{Standard Error $\sigma$= 2.102}\\
\end{tabular}
\end{table}

\begin{figure}
\centering
\includegraphics{Covid-19-Google-CMR-US_files/figure-latex/prediction-plots-1.pdf}
\caption{\label{fig:prediction-plots}Prediction surfaces at three points
during the pandemic according to the model; the dots are scatterplots of
the park- and residential-mobility indicators of the states on that
date.}
\end{figure}

The results suggest the potential of GCMR to investigate the spread of
COVID-19, but also point at some limitations. The baseline level is not
defined in a metric that is amenable to policy development (Google
defines ``mobility'' as an aggregated score of visits and length of stay
at places.) For example, it is not clear precisely what is residential
mobility: is it visits with friends and relatives, or mobility in the
vicinity of the place of residency? Without a clearer understanding of
these variables, their use can suggest trends, but their potential for
applied policy analysis appears to be more limited.

>>>>

  \begin{abstract}
  In 2020 Google released a set of Community Mobility Reports (GCMR).
  These reports are based on the company's location-tracking capabilities
  and measure changes in mobility with respect to a baseline. This novel
  source of data offers an opportunity to investigate potential
  correlations between mobility and transmission of COVID-19. Using data
  from the New York Times on COVID-19 cases and GCMR, this paper presents
  an analysis of mobility levels and new daily reported cases of COVID-19
  by state in the US. The results provide insights about the utility and
  interpretability of GCMR for COVID-19 research and decision-making.
  \end{abstract}

@@

This paper is a reproducible research document. The source is an
\texttt{R} markdown file available in a public
repository\footnote{See folder Covid-19-Google-CMR-US in \url{https://github.com/paezha/Google-Mobility-Reports-and-COVID-19-US/}}.

@@

GCMR use aggregated and anonymized data to chart changes in mobility
with respect to different classes of places (see Table
\ref{tab:descriptive-statistics}). Mobility indicators are calculated
based on the frequency and length of visits. The reports give percentage
change from a baseline level, which corresponds to the median value of
mobility of identical days of the week during the period between January
3 and Feb 6, 2020. Covid-19 data is compiled by The New York Times based
on reports from state and local health agencies.

\begin{table}[H]

\caption{\label{tab:descriptive-statistics}\label{tab:descriptive-statistics}Descriptive statistics of the data set}
\centering
\resizebox{\linewidth}{!}{
\fontsize{7}{9}\selectfont
\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{12em}llll}
\toprule
Variable & Definition & min & median & max & sd\\
\midrule
\rowcolor{gray!6}  New Cases & New Daily Cases of COVID-19 & 0 & 90 & 12312 & 1051.67\\
date & Date & 2020-02-27 & 2020-04-04 & 2020-05-02 & \\
\rowcolor{gray!6}  retail & Mobility trends for places like restaurants, cafes, shopping centers, theme parks, museums, libraries, and movie theaters & 0.34 & 0.66 & 1.16 & 0.21\\
groceries & Mobility trends for places like grocery markets, food warehouses, farmers markets, specialty food shops, drug stores, and pharmacies & 0.66 & 0.94 & 1.26 & 0.13\\
\rowcolor{gray!6}  parks & Mobility trends for places like local parks, national parks, public beaches, marinas, dog parks, plazas, and public gardens & 0.36 & 1.15 & 2.1 & 0.26\\
\addlinespace
transit & Mobility trends for places like public transport hubs such as subway, bus, and train stations & 0.24 & 0.72 & 1.14 & 0.23\\
\rowcolor{gray!6}  work & Mobility trends for places of work & 0.34 & 0.63 & 1.07 & 0.2\\
residential & Mobility trends for places of residence & 0.97 & 1.14 & 1.27 & 0.08\\
\bottomrule
\multicolumn{6}{l}{\textit{Note: }}\\
\multicolumn{6}{l}{All mobility indicators are lagged 11-day moving averages}\\
\end{tabular}}
\end{table}

For analysis, all mobility indicators are centered so that the value of
1 is the base mobility, and a 0.01 deviation corresponds to a 1\%
change. The incubation time of the disease is between 2 and 12 days
(95\% interval; see Lauer et al., 2020). Given this, it is to be
expected that any changes in mobility will have a lagged effect (if any)
on the discovery of new cases. For this reason, lagged moving averages
of the mobility indicators are calculated. Furthermore, it is possible
that mobility and reports of new cases of COVID-19 are endogenous, if
the public adjust their mobility according to reports of the incidence.
Therefore, in addition to being consistent with an incubation period,
use of lagged indicators also helps to break this potential endogeneity.

The lagged indicators are calculated as the mean of the mobility
indicator using the values from date-minus-12-days to date-minus-2-days.
Furthermore, using the cumulative number of reported COVID-19 cases, the
total number of new daily cases is calculated. This variable
(log-transformed after adding a small constant) is paired with the
corresponding lagged moving average of the mobility indicators. The
log-transformation is useful to avoid negative values of daily new cases
when making predictions. Table \ref{tab:descriptive-statistics} shows
the descriptive statistics of the data set. Analysis is based on
correlation analysis, multivariate regression, and data visualization.

@@

Table \ref{tab:correlation-analysis} shows that the mobility indicators
are highly correlated with each other. Two variables are selected for
multivariate analysis: parks- and work-related mobility. Work has a high
correlation with the outcome variable, and its correlation with parks is
relatively weak, which increases the information content of the two
variables. Furthermore, parks- and work-related mobility represent two
dimensions of out-of-home activities: mandatory and discretionary
travel.

\begin{table}[H]

\caption{\label{tab:check-correlations}\label{tab:correlation-analysis}Simple correlation between log(New Cases) and the mobility indicators}
\centering
\resizebox{\linewidth}{!}{
\fontsize{7}{9}\selectfont
\begin{tabular}[t]{lrrrrrrr}
\toprule
  & log\_new\_cases & retail & groceries & parks & transit & work & residential\\
\midrule
\rowcolor{gray!6}  log\_new\_cases & 1.00 & -0.68 & -0.48 & -0.27 & -0.68 & -0.69 & 0.71\\
retail & -0.68 & 1.00 & 0.87 & 0.41 & 0.93 & 0.98 & -0.98\\
\rowcolor{gray!6}  groceries & -0.48 & 0.87 & 1.00 & 0.44 & 0.87 & 0.87 & -0.87\\
parks & -0.27 & 0.41 & 0.44 & 1.00 & 0.49 & 0.39 & -0.41\\
\rowcolor{gray!6}  transit & -0.68 & 0.93 & 0.87 & 0.49 & 1.00 & 0.93 & -0.95\\
\addlinespace
work & -0.69 & 0.98 & 0.87 & 0.39 & 0.93 & 1.00 & -0.99\\
\rowcolor{gray!6}  residential & 0.71 & -0.98 & -0.87 & -0.41 & -0.95 & -0.99 & 1.00\\
\bottomrule
\multicolumn{8}{l}{\textit{Note: }}\\
\multicolumn{8}{l}{All mobility indicators are lagged 11-day moving averages}\\
\end{tabular}}
\end{table}

A regression model is estimated with the log of new daily cases as the
dependent variable. The covariates enter the regression in the form of a
second order polynomial expansion. In addition, the date (centered on
April 5) is introduced to account for the temporal trend of the
pandemic. Finally, an indicator variable for the state of New York and
an interaction with the date are used to distinguish the unusually high
incidence of the disease there. The results appear in Table
\ref{tab:model-results}. The model provides a good fit to the data and
all variables reported are significant at \(p<0.05\) or better.

There is an overall temporal trend that indicates a growing number of
cases over time, at an accelerating rate (see positive sign of
date\^{}2). Mobility related to parks and to work both tend to increase
the number of new cases; the negative sign for their interaction is
indicative of the trade-offs between these two forms of mobility and
their impact on new cases. The influence of parks-related mobility was
relatively weaker early in the pandemic (negative sign of the parks x
date term) but has become more important over time (positive sign of the
parks x date\^{}2 term). The opposite happens with work-related
mobility, which started with a greater effect (positive sign of work x
date term), but whose impact has declined over time (negative sign of
work x date\^{}2 term) New York has on average more new daily cases than
the rest of the states, but this has declined over time (see negative
sign of NY x date).

\begin{table}[H]

\caption{\label{tab:model-results}\label{tab:model-results}Results of estimating regression model. Dependent variable is log(New Cases + 0.0001).}
\centering
\fontsize{7}{9}\selectfont
\begin{tabular}[t]{lcc}
\toprule
Variable & Coefficient Estimate & p-value\\
\midrule
\rowcolor{gray!6}  date & 0.1085 & <0.001\\
date\textasciicircum{}2 & 0.0070 & <0.001\\
\rowcolor{gray!6}  parks\textasciicircum{}2 & 2.1756 & <0.001\\
parks & 8.9203 & <0.001\\
\rowcolor{gray!6}  parks x work & -23.6010 & <0.001\\
\addlinespace
work & 6.3156 & 0.0035\\
\rowcolor{gray!6}  work\textasciicircum{}2 & 12.0539 & <0.001\\
parks x date & -0.2133 & <0.001\\
\rowcolor{gray!6}  parks x date\textasciicircum{}2 & 0.0079 & <0.001\\
work x date & 0.1376 & 0.0272\\
\addlinespace
\rowcolor{gray!6}  work x date\textasciicircum{}2 & -0.0238 & <0.001\\
NY & 3.5466 & <0.001\\
\rowcolor{gray!6}  NY x date & -0.0527 & <0.001\\
\bottomrule
\multicolumn{3}{l}{\textit{Note: }}\\
\multicolumn{3}{l}{Coefficient of Determination $R^2$= 0.831}\\
\multicolumn{3}{l}{Adjusted Coefficient of Determination $R^2$= 0.83}\\
\multicolumn{3}{l}{Standard Error $\sigma$= 2.081}\\
\end{tabular}
\end{table}

Visualization is the most effective way to understand the trend
according to the mobility indicators and date. Figure
\ref{fig:prediction-plots} shows the prediction surfaces on three
different dates: March 21, when the first states began implementing
stay-at-home orders; April 5, fifteen days later; and April 20, fifteen
more days later, and at a time when some states started to consider
relaxing stay-at-home orders.

On March 21 there were still only minor departures from baseline
mobility (recall that these are temporally lagged); the prediction
surface at this point is relatively flat. This changes by April 5, when
work-based mobility has declined substantially. Although every state
registers lower work-based mobility, there are large variations in
parks-based mobility, with some states seeing increases of up to 60\%
for this class of mobility.

The prediction surface indicates an expectation of a greater number of
new daily cases as either class of mobility increases, but the effect is
not linear. On the last date examined, April 20, the trend becomes more
steep for park-based mobility, even as this indicator continues to
display large variations from the baseline in both directions. The white
dashed lines in the plots are the folds of the saddles, and represent,
for each date, the combination of parks- and work-related mobility
levels that tended to minimize the emergence of new cases.

\begin{figure}
\centering
\includegraphics{Covid-19-Google-CMR-US_files/figure-latex/prediction-plots-1.pdf}
\caption{\label{fig:prediction-plots}Prediction surfaces at three points
during the pandemic according to the model; the dots are a scatterplot
of the parks- and work-related mobility indicators of the states on that
date; the white dashed line is the fold of the saddle.}
\end{figure}

The results suggest that over time the benefits of reduced work-related
mobility can be offset by parks-related mobility. For example, on March
21, New Jersey and Idaho had similar levels of park-related mobility. By
April 20, New Jersey had substantially reduced park-related mobility,
whereas Idaho's was even higher than in March. The estimated and actual
number of new cases grew in the intervening period; however, New
Jersey's growth in new cases (from March 21 to April 20) was only 894\%,
whereas Idaho's was 1030\%.

These results suggest the potential of GCMR to investigate the spread of
COVID-19, but also point at some limitations. The baseline level is not
defined in a metric that is amenable to policy development (e.g.,
person-km travelled). Without a clearer understanding of the absolute
levels of these variables, their use can suggest trends, but their
potential for applied policy analysis appears to be more limited.
